---
title: "Political consumerism: undertstanding individual propensity to boycott"
author: "Chiara Pievani - ID 84347A"
format: html
editor: visual
bibliography: references.bib
---

```{r setup}
#| include: false

library(tidyverse)
library(haven)
library(here)
library(lme4)
library(sjPlot)
library(marginaleffects)
options(marginaleffects_safe = FALSE)

```

::: {style="text-align: center;"}
*What factors are associated with individual’s likelihood of boycotting products as a form of political dissent across European countries?*
:::

Consumer boycotts are the refusal to purchase from specific companies to achieve social or political aims (Friedman, 1999).

This research wants to investigate which factors are associated with participation in product boycotts across European countries using data from the European Social Survey Round 11 Multilevel dataset. Building on previous research, this study focuses on the relationship between boycott participation and political engagement, support of the respondent for their own country's institutions, political orientation and concern about climate change. Additionally, the analysis explores whether the association between boycott participation and climate concern varies across the political orientation scale.

The following hypotheses regarding the individual-level factors associated with participation in boycotts are then tested:

H1: Individuals with higher levels of political engagement are more likely to consider boycott as a form of political participation.

H2: Respondents who show greater support for their country's institutions are less likely to participate in boycotts.

H3: Individuals who are more concerned with climate change will be more likely to participate in boycotts.

H4: Climate concern is expected to have a higher association with boycotting among people that, on the ideology scale, place themselves on the left.

H5: Younger and more educated respondents are more likely to engage in product boycotts.

H6: People living in urban areas are more likely to engage in political consumerism practices like boycotts.

## Data and variable operationalization

The dataset used is the ESS Round 11 Multilevel dataset. In its original format, it contains over forty-six thousand observations across around nine thousand three hundred variables that collect data both at the individual and regional, national level.

To avoid any kind of extrapolation, only complete observations have been considered for this analysis. Considering it is a large dataset, this decision still left me with over thirty thousand respondents to consider.

To try better understand the relationship I have described in the previous section, I have selected fifteen of the variables included in the dataset.

The dependent variable is "Boycotted certain products in the last 12 months" (*bctprd*): it is a binary value (No = 0, Yes = 1).

With the independent variables chosen, I want to investigate how both political ideology and concern for climate change are related with the propensity to boycott.

The independent variables considered are then:

- Political interest (*polintr*), measured on a scale 1 - 4; for interpretation purposes, the scale has been reversed, so that 1 = "Not at all interested" and 4 = "Very interested;

- Taken part in a public demonstration in the last 12 months (*pbldmna*), which is binary and 0 = No, 1 = Yes;

- Trust in politicians (*trstplt*) and Trust in parliament (*trstprl*), both measured on a scale 0 - 10, in which 0 = "No trust at all" and 10 = "Complete trust";

- Satisfaction with national government (*stfgov*), measured on a 0 - 10 scale, in which 0 = "Extremely dissatisfied" and 10 = "Extremely satisfied";

- Self placement on left - right scale (*lrscale*) measured on a 0 - 10 scale. Compared to the original variable in the ESS dataset, in the new variable *lrscale_c*, the scale has been reversed, so that 0 = "Extreme right" and 10 = "Extreme left", and then mean centred based on the new reversed scaling (mean = 5.04): this has been done to facilitate and make more straightforward the interpretation of the interaction stated in the previous section under H3, so that the value 0 now indicates placement in the centre of the ideological scale;

- Worry about climate change (*wrclmch*) measured on a scale from 1 - 5 and 1 = "Not at all worried", while 5 = "Extremely worried";

- Personal responsibility to reduce climate change (*ccrdprs*), measured on a scale from 0 - 10, in which 0 = "Not at all" and 10 = "A great deal".

Other relevant factors included in the model used as control variables to further investigate the relationship between the dependent and the independent variables are:

- Age (*agea*), continuous;

- Years of education (*eduyrs*), continuous;

- Gender (*gndr*), binary where 0 = "Male" and 1 = "Female"

- Residence (*rsdnc*), binary where 0 = "Rural residence" and 1 = "Urban residence". This variable is a modification of the variable *domicil* of the ESS dataset: those who said to live in either a big city, the suburbs of a big city or a town have been assigned to the category "Urban residence", while those who said to live in a country village or a home in the countryside have been assigned to the category "Rural residence".

- Country (*cntry*), used as a clustering variable to see if the hypothesized relationship changes across different countries.

All continuous variables have been standardized.

### Operationalization

Some of the variables chosen for the analysis capture concepts close to one another. To simplify the interpretation of the model, these variables were operationalized into additive indices.

Worry about climate change (*wrclmch*) and personal responsibility to reduce climate change (*ccrdprs*) presented a correlation of 0.46, which I considered enough to say that these two variables could be measuring a common latent dimension, that I called "Climate concern" (*clmcnrn*). To build the index, the two original variables were first standardized, since they were measured on different scales, and then averaged.

Trust in politicians (*trstplt*), trust in parliament (*trstprl*) and satisfaction with national government (*stfgov*) present correlations that range between 0.58 and 0.75, high enough to suggest that the three could be explaining a common latent factor. To confirm this, Eigen values have been calculated, and they were coherent with the conclusion taken by looking at the correlation plot, suggesting that only one factor should be retained to explain these three variables. Lastly, also in this case I built an additive index of the three, called "Political support" (*pltsprt*)

```{r processed data import}
#| include: false

data <- read.csv(here("data", "processed", "clean_data.CSV"))

# for full code on data cleaning see script 01_data_prep.R

```

## Model choice and statistical justification

Since the dependent variable (*bctprd*) is binary, a linear regression model would not be coherent. To ensure that the predicted probabilities stay bounded between 0 and 1, the model to choose is the logistic one.

Although, to answer the research question, a multivariate logistic regression model might not be good enough to unveil the full relationship between boycott, politics and environmental concern.

The ESS dataset has a hierarchical structure, hence it is possible to not only consider individuals as single units but also to consider them as nested within countries. This because people living in the same country share the same institutional and cultural background, and in cases like political dissent, this common context is likely to influence their attitude towards it.

To assess whether the clustering should be taken into account to better explain the relationship described previously, an explorative empty logistic multilevel model with random intercepts for countries was estimated. This allows to see the Intra-Class Correlation, hence the variation explained at the group level.

```{r empty model}
#| echo: false

empty_model <- glmer(
  bctprd ~ 1 + (1 | cntry),
  data = data,
  family = binomial(link = "logit")
)

table_mod0 <- tab_model(empty_model,
                  title = "Table 1. Empty logistic multilevel model predicting boycott participation")

table_mod0

```

The table above reports an ICC of 18%, which is the portion of variance explained by country-level differences.

This confirms that not considering the differences brought by countries to the model would affect the validity of the results.

Eventually, the relationship that this analysis will attempt to look further into is:

`bctprd  ~ polintr + pbldmna + pltsprt + lrscale * clmcnrn + (1 + clmcnrn|cntry) + z_age + z_edu + gndr +  rsdnc`

This includes then indicators for political engagement and orientation and environmental concern. An interaction term between ideological self-placement and climate concern is included to see whether the relationship between environmental concern and participation in boycotts changes across the left - right scale.

Moreover, both the baseline and the association between climate concern and participation in boycotts are allowed to vary across countries in the multilevel term of the model.

Age, education, gender and place of residence are included as control variables.

## Results and interpretation

Since there are many predictors to consider, for better interpretation purposes and better assessment of the contribution of different sets of variables, the analysis develops incrementally.

Politics and climate related variables are first considered separately to see the contributions of each context.

Then, a full model with all explanatory variables is estimated, with all control variables included.

Finally, the interaction between ideological placement and climate concern and the random slope for climate concern are included to explore whether the effect of environmental attitudes varies across both individuals (based on their political orientation) and countries.

### Model 1: separate model for predictors with random intercepts for countries

The first step of the analysis is then to estimate logistic multilevel models for political and environmental predictors, keeping them separate, in order to evaluate in a first moment their independent association with boycott participation.

```{r model 1a and 1b}
#| echo: false

# Model 1 - separate predictors ----------------------------------------

model_1a <- glmer(
  bctprd ~ polintr + pbldmna + pltsprt + lrscale_c + 
  (1 | cntry), data = data,
  family = binomial(link = "logit")
  )

model_1b <- glmer(
  bctprd ~ clmcnrn + (1 | cntry),
  data = data, family = binomial(link = "logit")
)

table_mod1 <- tab_model(model_1a, model_1b,
          transform = NULL,
          title = "Table 2. Regression coefficients for political and environmental predictors with random intercepts for countries")

table_mod1

```

Already from this first models, we can see that political engagement is positively associated with participation in a product boycott.

In particular, the relationship is stronger between respondents who participate in public demonstrations and those that take part in product boycotts. This means that, holding all other predictors constant, participation in demonstrations is associated with a maximum increase of about 33% in the probability of boycotting a product.

This very strong association can be brought back to the fact that both boycotts and demonstrations are unconventional forms of political participation.

Regarding climate concern, individuals who express greater concern about climate change are more likely to say that they participated in a product boycott. This is consistent with the idea that product boycott is associated not only with political engagement, but also with ethical consumption.

Finally, for this model, the ICC confirms the importance of implementing a multilevel model: individuals with similar characteristics might have different probabilities of participating in a boycott depending on the country they live in.

### Model 2: full individual model with control variables

While model 1 estimated participation in product boycotts keeping the political and environmental predictors separate, model 2 combines them into a single multilevel logistic regression model, including also control variables for age, education, gender and place of residence.

This way it is now possible to interpret the coefficients as conditional to all the other variables, which are held constant, and then interpret them as independent effects.

Now it is possible to see whether, among two respondents who show the same level on all the political predictors, the individual who is more concerned about climate change is more likely to participate in a boycott, and viceversa.

```{r model 2}
#| echo: false

# Model 2 - all predictors + controls --------------------------

model_2 <- glmer(
  bctprd ~ polintr + pbldmna + pltsprt + lrscale_c + 
  clmcnrn + (1 | cntry) +
  z_age + z_edu + gndr + rsdnc, 
  data = data, family = binomial(link = "logit")
)

table_mod2 <- tab_model(model_2,
          transform = NULL,
          title = "Table 3. Regression coefficients for all individual predictors with random intercepts for countries")

table_mod2

```

Indeed, holding all other predictors constant and comparing respondents from the same country, a one-unit increase in political interest is associated with a maximum increase in the probability of participating in a boycott of approximately 8.5%.

Moreover, with this model we can see that holding all political predictors and control variables constant, a one-unit increase in climate concern is associated with a maximum increase in the probability of participating in a boycott of about 9.7%.

As far as the control variables are concerned, age has a negative coefficient, indicating that younger individuals might be more inclined to participate in a product boycott; same goes for respondents who have spent more years in education and for women (*gndr* = 1). Place of residence is not statistically significant, hence I cannot conclude that there is a significant difference among respondents from rural places or urban ones in their propensity to boycott.

### Model 3: full multilevel logistic regression model with random slopes and intercepts and interaction term

The final model builds on the previous one and it adds two fundamental items for further understanding of the main independent variables of the model.

The first one is an interaction term between climate concern and the centered left - right ideological scale, while the second one is the term that allows both the intercepts and the slopes for climate concern to vary across countries.

These two elements make it possible to see whether climate concern association with participation in boycotts differs based on respondents' ideological orientation and across different countries.

```{r model 3}
#| echo: false

# Model 3 - random slopes and interaction --------------------

model_3 <- glmer(
  bctprd ~ polintr + pbldmna + pltsprt + lrscale_c + 
  clmcnrn + (1 + clmcnrn | cntry) + lrscale_c:clmcnrn +
  z_age + z_edu + gndr + rsdnc, 
  data = data, family = binomial(link = "logit")
)

table_mod3 <- tab_model(model_3,
                        transform = NULL,
                        title = "Table 4. Regression coefficients for full multilevel logistic regression model with random slopes and intercepts for climate concern across countries and interaction between climate concern and ideological placement")

table_mod3

```

As said in the section about data description and operationalization, the political orientation scale was reversed and mean centered, so that the value 0 now represents the center of the scale, while values below 0 represent right-wing positions and values above 0 left-wing positions.

With the interaction term, the coefficient for climate concern (which remained positive and statistically significant), now represents the associated increase with the probability in participating in a boycott for people who place themselves at the average of the ideological scale, which for this variable was originally around 5.

The coefficient of the interaction between climate concern and political orientation is also positive and statistically significant. This indicates that the association of climate concern with the dependent variable becomes stronger as individuals move towards higher values of the scale, hence as they are leaning more towards left ideology.

```{r predicted probabilities plot}
#| echo: false

clmcnrn_lrscale <- plot_slopes(
  model_3,
  variables = "clmcnrn",
  condition = "lrscale_c"
) +
  geom_rug(
    data = data,
    aes (x = lrscale_c),
    inherit.aes = FALSE,
    sides = "Slope",
    alpha = 0.2
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color ="grey40")+
  theme_bw(base_size=10) +
  labs(title = "Figure 1. Average Marginal Effects plot of clmcnrn on bctprd across lrscale")

clmcnrn_lrscale

```

The marginal effects plot above clearly shows that the association between climate concern and boycott participation is positive for both right-wing and left-wing respondents, but it becomes stronger as individuals are politically oriented more towards left.

The coefficients for all the other predictors overall remain unchanged compared to the previous model.

Regarding the random effects section of the regression table, the variance of the random slope for climate concern (0.04, SD = 0.2) indicates that the effect of climate concern is not constant across countries, meaning that climate concern translates into political consumerism more strongly in some countries compared to others.

Lastly, the positive correlation between random intercepts and random slopes (0.66) indicates that countries with a higher intercept value also tend to have a stronger association between climate concern and participation in boycotts.

In countries where climate concern is a more heartfelt issue and might already have been institutionalized with green policies, then these countries' citizens might see boycott as a legitimate translation to address these issues.

## Substantive insights

The results obtained support most of the hypotheses proposed in this report.

Political engagement is one of the most important factors associated with boycott participation; individuals who report higher levels of political interest and, in particular, those who have participated in demonstrations are significantly more likely to engage in product boycotts. Boycotting then is not a simple consumer choice, but rather a form of political participation used by individuals who are already politically active.

Secondly, findings are consistent with the theory that individuals who think that their country's institutions are trustworthy might feel less the need of using boycotts to express their political opinions.

Strong evidence also supports the third hypothesis. Considering all the other independent variables and the socio-demographic controls, climate concern stays positively associated with boycott participation. Environmental awareness then could be considered as an independent motivation for political consumerism.

Moreover, there is a positive and significant interaction between climate concern and political orientation, suggesting that the association between climate concern and boycott participation becomes stronger among left-wing respondents.

Regarding socio-demographic control variables, results only support H5, regarding younger and more educated respondents. About H6, statistical significance is too weak, therefore it is not possible to conclude that there is a sufficiently strong association between living in an urban area and propensity to participate in boycotts.

## Limitations and reflection

The analysis is based on cross-sectional data: this does not allow for any causal interpretation of the relationships found in the models. For instance, individuals might start participating in boycotts and only in a second time become more aware and concerned with climate change issues, rather than the other way around.

Moreover, no country-level predictors were included in the analysis, although the models account for variation within countries. Therefore, the observed differences among countries remain mostly unexplained. Predictors that could be included in this analysis could be GDP per capita or country-level indicators of environmental performance.

One last limitation of this project is the possibility of having other factors that are plausibly related to boycott participation, for instance social trust, media consumption or variables that measure general attitudes towards ethical consumption.

All things considered, the most important finding in my opinion is the role played by political orientation in shaping the relationship between climate concern and boycott participation, shown by the interaction in the final model.

## Bibliography

[@baptista2024; @consumer1999; @koos2012; @mata2023; @neilson2010]

## AI disclosure

Artificial Intelligence has been used as a coding support to:

- Reverse the scaling of some of the variables;

- Compute the eigenvalues for the operationalization;

- Solve issues with the saving of the processed data file;

- Show log-odds coefficients instead of odd-ratios in the regression tables;

- Commands for better outcomes in the rendered file;

- Help with linguistic revision.
